---
title: "Avian Speciation"
author:
date: | 
  | Office: 305 Lewis Hall
  | Dr. Ethan Linck
output: beamer_presentation
fontsize: 11pt
theme: boxes
header-includes:
  - \addtobeamertemplate{title page}{\centering \includegraphics[width=1\textwidth]{images/banner2.png}}{}
knit: (function(input, ...) {
    out <- rmarkdown::render(input, ...);
    tools::compactPDF(out, gs_quality = "printer");
    out
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# What is a species?

\begin{figure}
\includegraphics[width = \linewidth]{images/flycatchers.png}
\end{figure}  

\hfill \tiny ([*Empidonax difficilis*](https://macaulaylibrary.org/asset/236799211) / [*Empidonax occidentalis*](https://macaulaylibrary.org/asset/108268701))  

# Species concepts

- 24 species concepts and counting (Mayden 1997)
- Widespread confusion between conceptual basis and operational criteria (*definitions* versus *delimitation*)
- Conflict from alternate definitions, insufficient data...

\begin{figure}
\includegraphics[width = 0.4\linewidth]{images/phylloscopus.png}
\end{figure}  

# Species concepts

\begin{figure}
\includegraphics[width = \linewidth]{images/dequeiroz.png}
\end{figure}  
(De Queiroz 2007)

# Species concepts  

- *Phylogenetic species concept*: A species is a group of individuals that conform to one or more phylogenetic criteria (reciprocal monophyly, coalescence of all gene trees)

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/phylogeny.png}
\end{figure}  

# Species concepts  

- *Ecological species concept*: A species is a group of individuals that share the same ecological niche or adaptive zone

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/nicheconcept.png}
\end{figure}  

# Species concepts  

- *Evolutionary species concept*: A species is a group of individuals with the same evolutionary role, tendencies, and historical fate

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/evolutionary.png}
\end{figure}  

# Species concepts  

- *Phenetic cluster species concept*: A species is a group of individuals that form a cluster in morphological space separated from other such groups (similar trait values for multiple features)

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/phenetic.png}
\end{figure}  

# Species concepts 

- *Genotypic cluster species concept*: A species is a group of individuals that form a cluster in genoptypic space separated from other such groups (similar allele frequencies at multiple loci)

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/genotypic.png}
\end{figure}  


# Species concepts

- *Biological species concept*: A species is a group of actively or potentially interbeeding individuals separated from other such groups by intrinsic or extrinsic reproductive isolation (not only geography)  

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/biological.png}
\end{figure}  

# Speciation is a spectrum

- Another source of conflict: some taxa caught in middle of the process!

\begin{figure}
\includegraphics[width = 0.45\linewidth]{images/dequeiroz2.png}
\end{figure}  
\hfill \tiny  (De Queiroz 2007)

# Which of the following is a *criterion* for delimiting species? 

a) species have genomes that are greater than 5% different than their closest relative
b) species are groups of individuals that share an ecological niche
c) species are members of a lineage with a shared evolutionary trajectory
e) none of the above
f) all of the above

# Prezygotic Isolation

- Divergence *before* mating: ecology, behavior, appearance, song
- Can be measured with field experiments to test for intraspecific aggression against potential competitors for mates

\begin{figure}
\includegraphics[width = 0.8\linewidth]{images/freemansong.png}
\end{figure}  

\hfill \tiny (Freeman et al. 2022)  

# Prezygotic Isolation

- Freeman et al. 2020: birds that do *not* respond aggressively to songs from close relatives ("song discrimination") in allopatry (isolation) are less likely to interbreed in parapatry (i.e., contact)   
- Birds with the same level of genetic divergence but greater song discrimintation are less likely to interbreed in parapatry  

\begin{figure}
\includegraphics[width = 0.8\linewidth]{images/freemansong.png}
\end{figure}  

\hfill \tiny (Freeman et al. 2022)  

# Postzygotic Isolation 

- Divergence *after* mating: genetic incompatabilities, selection against hybrids
- Can be measured by fitness or prevalance of hybrids, experiments, assessment of genetic architecture 

\begin{figure}
\includegraphics[width = 0.8\linewidth]{images/postzygotic.png}
\end{figure}  

\hfill \tiny (Price et al. 2002)  

# Allopatric speciation

- Speciation in geographic isolation ("allo"=different, "patry"=homeland)
- Predicts allopatric populations are monophyletic groups

\begin{figure}
\includegraphics[width = \linewidth]{images/juncos.png}
\end{figure}  

\hfill \tiny (Freis et al. 2016)  

# Allopatric speciation

- "Easiest" and by far most common mode
- Allopatric ranges can be generated by geologic processes, climatic changes, dispersal 

\begin{figure}
\includegraphics[width = \linewidth]{images/allopatric.png}
\end{figure}  

\hfill \tiny (Kearns et al. 2014)  

# Is sympatric speciation possible?

- Speciation *without* geographic barriers to gene flow
- Requires strong natural selection 

\begin{figure}
\includegraphics[width = \linewidth]{images/sympatric1.png}
\end{figure}  

\hfill \tiny (Ryan et al. 2016)  

# Which prediction is consistent with sympatric speciation?

- Tips with the same number are on the same island
- Tips with the same letter have the same ecological adaptation

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.height=3.25, fig.width=5}
library(ape)

# Hypothesis 1: Parallel sympatric speciation
tree_sympatric <- read.tree(text = "((A1:1,B1:1):1,(A2:1,B2:1):1);")

# Hypothesis 2: Allopatric speciation
tree_allopatric <- read.tree(text = "((A1:1,A2:1):1,(B1:1,B2:1):1);")

# Plot side-by-side
par(mfrow = c(1, 2))
plot(tree_sympatric, main = "a)", cex = 1.2)
plot(tree_allopatric, main = "b)", cex = 1.2)
```

(in **a**, birds on the same island have different niches; in **b**, birds on different islands have the same niche)  

# Sympatric speciation in South Atlantic Finches

- Bill size matches seeds in habitat 

\begin{figure}
\includegraphics[width = \linewidth]{images/sympatric2.png}
\end{figure}  

\hfill \tiny (Ryan et al. 2016)  

# Sympatric speciation in South Atlantic Finches

- Different ecological forms within each island are most closely related

\begin{figure}
\includegraphics[width = 0.7\linewidth]{images/sympatric3.png}
\end{figure}  

\hfill \tiny (Ryan et al. 2016)  

# Parapatric speciation

- Intermediate to allopatric and sympatric speciation
- Adjacent but nonoverlapping ranges

\begin{figure}
\includegraphics[width = 0.7\linewidth]{images/syma.png}
\end{figure}  
\hfill \tiny (Linck et al. 2020, *J. Evol. Biol.*)

\normalsize
(in Linck et al., Mountain Kingfisher diverged from Yellow-billed Kingfisher once, not repeatedly across elevational gradients, making paraptric speciation hard to prove, though they do hybridize)  

# Ring species

- Adjacent populations similar but subtly different
- Reproductive isolation increases with distance, complete at ends of horseshoe
- Example: Greenish Warbler wrapping around the Tibetan Plateau; morphologically and ecologically similar. 

\begin{figure}
\includegraphics[width = 0.9\linewidth]{images/ringspecies.png}
\end{figure}  

\hfill \tiny (Irwin et al. 2001; [*Phylloscopus trochiloides*](https://macaulaylibrary.org/asset/44704121))  

# Would the Biological Species Concept split Greenish Warbler?

\begin{figure}
\includegraphics[width = 0.6\linewidth]{images/ringspecies.png}
\end{figure}  

\hfill \tiny (Irwin et al. 2001; [*Phylloscopus trochiloides*](https://macaulaylibrary.org/asset/44704121))  

\normalsize

a) Yes
b) No
c) It depends on geographic sampling 


# Lecture Exam Review

- Wednesday, September 17th
- 50 points on avian diversity, taxonomy, biogeography, phylogenetics, and speciation
- No lab content that wasn't also covered in lecture (e.g., what a hallux is )
- Question types: multiple choice, multiple select, true / false, fill-in-the-blanks, short answer
- Format: Canvas quiz, closed book. Bring a laptop or tablet. 
- **Anything covered in lecture is fair game**

# Diversity

- Know what an apomorphy is
- Know what traits define birds
- Be able to qualitatively describe how the number of described species has changed through time, and why
- Know what an adaptive radiation is
- Know how bird bodies, feet, bills, habitat use reflects the process of adaptive radiation 

# Taxonomy & Biogeography

- What taxonomic class do birds belong to?
- Know how many avian classes, orders, families, genera, and species there are to an order of magnitude
- Be able to qualitatively describe how the number of described species has changed through time, and why
- What latin suffix does each avian order have? Each family?
- Know at least one taxon (order, family, genus, or species) from each higher level group in the `03_taxonomy_biogeography.pdf` lecture
- Know at least one taxon associated with each biogeographic realm
- Know what "endemism" means, and what causes it

# Evolutionary origins

- Why are birds considered reptiles in a phylogenetic sense?
- What traits do birds share with reptiles?
- Did hollow bones and feathers follow the evolution of flight, or vice versa?
- Modern bird traits present and absent in *Archaeopteryx*
- Bird-hipped versus reptile-hipped dinosaurs
- Cursorial versus arboreal hypotheses for the origin of flight
- K-Pg boundary, Mesozoic versus Cenozoic 
- Cladogenesis versus anagenesis 

# Phylogenetics and Systematics

- Terminology: monophyly, paraphyly, apomorphy, polytomy
- Skills: Reading trees, identifying monophyletic groups
- Knowledge: describe / name members of Australaves, Paleognathae, Neoaves, Galliformes, Anseriformes, Aequiornithes, "Basal" Neoaves
- Remember: how taxonomy and phylogenetics are related, why relationships basal Neoaves are controversial
- What are the limitations of morphological data, and what are the advantages of DNA? 
- Why do recent higher-level avian phylogenies conflict with each other? 
- Why might two species be "lumped"?

# Species concepts and speciation

- Be able to define / identify major species concept categories (e.g., phylogenetic versus biological versus ecological)
- Prezygotic versus postzygotic isolation (description and examples)
- Allopatric versus sympatric verys and parapatric speciation
- Which mode is most common, and how do we know? 
- What is a ring species? 

# References:

- Title Slide Species: *Trogon citreolus*, *Pitta sordida*, *Icterus icterus*, *Pachycephala macrorhyncha*

- Kearns, A. M., Joseph, L., Toon, A., & Cook, L. G. (2014). Australia’s arid-adapted butcherbirds experienced range expansions during Pleistocene glacial maxima. Nature communications, 5(1), 3994.

- Freeman, B. G., Rolland, J., Montgomery, G. A., & Schluter, D. (2022). Faster evolution of a premating reproductive barrier is not associated with faster speciation rates in New World passerine birds. Proceedings of the Royal Society B, 289(1966), 20211514.

- Price, T. D., & Bouvier, M. M. (2002). The evolution of F1 postzygotic incompatibilities in birds. Evolution, 56(10), 2083-2089.

- Irwin, D. E., Bensch, S., & Price, T. D. (2001). Speciation in a ring. Nature, 409(6818), 333-337.

# References:

- Linck, E., Freeman, B. G., & Dumbacher, J. P. (2020). Speciation and gene flow across an elevational gradient in New Guinea kingfishers. Journal of Evolutionary Biology, 33(11), 1643-1652.

- Ryan, P. G., Bloomer, P., Moloney, C. L., Grant, T. J., & Delport, W. (2007). Ecological speciation in South Atlantic island finches. Science, 315(5817), 1420-1423.
