---
title: "Modern Bird Diversity"
author:
date: | 
  | Office: 305 Lewis Hall
  | Dr. Ethan Linck
output: beamer_presentation
fontsize: 11pt
theme: boxes
header-includes:
  - \addtobeamertemplate{title page}{\centering \includegraphics[width=1\textwidth]{images/banner2.png}}{}
knit: (function(input, ...) {
    out <- rmarkdown::render(input, ...);
    tools::compactPDF(out, gs_quality = "printer");
    out
  })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Classifying Modern Birds

- Class: Aves
- Order: Passeriformes
- Family: Turdidae
- Genus: *Ixoreus*
- Species: *I. naevius*
- Specific Epithet: *naevius*


\begin{figure}
\includegraphics[width = 0.5\linewidth]{images/vath.png}
\end{figure}  
\hfill \tiny ([ML28703721](https://macaulaylibrary.org/asset/28703721))

# Classifying Modern Birds

How do we quantify taxonomic diversity in birds?

- 1 class
- ~40 orders
- ~247 families
- ~2312 genera
- ~10,699 species 

*via Gill & Prum. Know these to an order of magnitude!

# Why might species numbers change over time?

```{r, message=FALSE, warning=FALSE, echo=FALSE, fig.width=5, fig.height=3, out.width="95%"}
library(ggplot2)
birds <-c(18939, 8590, 9672, 10425, 11145)
years <- c(1909, 1951, 1990, 2014, 2024)
source <- c("Sharpe", "Mayr & Amadon", "Sibley & Monroe", "eBird 2014", "eBird 2025")
df <- cbind.data.frame(birds, years, source)
ggplot(df, aes(x = years, y = birds, shape = source)) +
  geom_line(aes(group = 1), linetype = "dashed", size = 0.4, color = "black") +
  geom_point(size = 3) +
  theme_classic() +
  theme(legend.position = c(1, 1),         # upper right inside plot
        legend.justification = c(1, 1)) +
  labs(x = "Year", y = "Number of species", shape = "Source")

```

# Higher-level Avian Taxonomy

\begin{figure}
\includegraphics[width = \linewidth]{images/birdtree.png}
\end{figure}  
\hfill \tiny (Stiller et al. 2024)

# Paleognathae

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/coos.png}
\end{figure}  

Stuthioniformes (ostriches, 2 spp.)

\hfill \tiny ([*Struthio camelus*](https://macaulaylibrary.org/asset/204565301))

# Paleognathae  

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/rhea.png}
\end{figure}  

Rheiformes (Rheas, 2 spp.)

\hfill \tiny ([*Rhea americana*](https://macaulaylibrary.org/asset/72353001))

# Paleognathae  

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/kiwi.png}
\end{figure}  

Apterygiformes (Kiwis, 5 spp.) 

\hfill \tiny ([*Apteryx maxima*](https://macaulaylibrary.org/asset/219724521))

# Paleognathae  

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/cassowary.png}
\end{figure}  

Casuariformes (Emus and Cassowaries, 4 spp.)

\hfill \tiny ([*Casuarius casuarius*](https://macaulaylibrary.org/asset/63994411))

# Paleognathae  

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/tinamou.png}
\end{figure}  

Tinamiformes (Tinamous, 47 spp.)

\hfill \tiny ([*Tinamus major*](https://ebird.org/checklist/S52561962))

# Neognathae: Galloanseres

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/brushturkey.png}
\end{figure}  

Important orders: Galliformes ("landfowl", 300 spp.)

\hfill \tiny ([*Alectura lathami*](https://macaulaylibrary.org/asset/81277181))

# Neognathae: Galloanseres

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/magpiegoose.png}
\end{figure}  

Anseriformes ("waterfowl", 177 spp.)

\hfill \tiny ([*Anseranas semipalmata*](https://macaulaylibrary.org/asset/70134081))

# Neognathae: "Orphaned" Neoaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/nightjar.png}
\end{figure}  

Caprimulgiformes (swifts, nightjars, hummingbirds, etc., 602 spp.)

\hfill \tiny ([*Uropsalis lyra*](https://macaulaylibrary.org/asset/51940921))

# Neognathae: "Orphaned" Neoaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/cuckoo.png}
\end{figure}  

Cuculiformaces (cuckoos, 149 spp.)

\hfill \tiny ([*Coccyzus longirostris*](https://macaulaylibrary.org/asset/80877841))

# Neognathae: "Orphaned" Neoaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/ptilinopus.png}
\end{figure}  

Columbiformes (pigeons, 344 spp.)

\hfill \tiny ([*Ptilinopus magnificus*](https://macaulaylibrary.org/asset/246127211))

# Neognathae: "Orphaned" Neoaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/franklin.png}
\end{figure}  

Charadriiformes (shorebirds, gulls, terns, etc., 383 spp.)

\hfill \tiny ([*Leucophaeus pipixcan*](https://macaulaylibrary.org/asset/55259551))

# Neognathae: Neoaves: Aequornithes

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/penguin.png}
\end{figure}  

Sphenisciformes (penguins! 18 spp.)

\hfill \tiny ([*Pygoscelis papua*](https://macaulaylibrary.org/asset/612758547))

# Neognathae: Neoaves: Aequornithes

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/albatross.png}
\end{figure}  

Procelariiformes (tube-nosed seabirds, e.g. petrels and albatrosses, 147 spp.)

\hfill \tiny ([*Diomedea sanfordi*](https://macaulaylibrary.org/asset/262994551))

# Neognathae: Neoaves: Aequornithes

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/ibis.png}
\end{figure}  

Pelecaniformes (pelicans, ibises, Shoebill, etc., 118 spp.)

\hfill \tiny ([*Theristicus caudatus*](https://macaulaylibrary.org/asset/44739891))

# Neognathae: Neoaves: Telluraves: Afroaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/astur.png}
\end{figure}  

Accipitriformes (raptors, 266 spp.) 

\hfill \tiny ([*Theristicus caudatus*](https://macaulaylibrary.org/asset/298044011))

# Neognathae: Neoaves: Telluraves: Afroaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/owl.png}
\end{figure}  

Strigiformes (owls, 244 spp.)

\hfill \tiny ([*Otus rutilus*](https://macaulaylibrary.org/asset/290921711))

# Neognathae: Neoaves: Telluraves: Afroaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/motmot.png}
\end{figure}  

Coraciiformes (kingfishers and relatives, 177 spp.)

\hfill \tiny ([*Baryphthengus martii*](https://macaulaylibrary.org/asset/115091291))

# Neognathae: Neoaves: Telluraves: Afroaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/hemicircus.png}
\end{figure}  

Piciformes (woodpeckers and relatives, 445 spp.) 

\hfill \tiny ([*Hemicircus concretus*](https://macaulaylibrary.org/asset/122514161))

# Neognathae: Neoaves: Telluraves: Australaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/palm.png}
\end{figure}  

Psittaciformes (parrots, 398 spp.)

\hfill \tiny ([*Probosciger aterrimus*](https://macaulaylibrary.org/asset/61124271)

# Neognathae: Neoaves: Telluraves: Australaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/falcon.png}
\end{figure}  

Falconiformes (falcons / "murder parrots", 67 spp.)

\hfill \tiny ([*Herpetotheres cachinnans*](https://macaulaylibrary.org/asset/84034321))

# Neognathae: Neoaves: Telluraves: Australaves

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/townsend.png}
\end{figure}  

Passeriformes ("perching birds", 6,456 spp.) (!)

\hfill \tiny ([*Setophaga townsendi*](https://macaulaylibrary.org/asset/45702491))

# Which of the following is *not* a member of Neoaves?

a) Canada Goose  
b) Ring-necked Parakeet    
c) Rufous Hummingbird  
d) Ring-billed Gull  
e) Burrowing Owl  

# Biogeographic Realms

\begin{figure}
\includegraphics[width = \linewidth]{images/realms.png}
\end{figure}  

\hfill \tiny ([*Herpetotheres cachinnans*](Udvardy 1975))

# Which phylogeny is more consistent with these regions?

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=5, fig.height=3, out.width="95%"}
library(ape)

set.seed(917)

# helper for colors
get_colors <- function(labels) {
  ifelse(grepl("Nearctic", labels), "steelblue", "firebrick")
}

# --- Case 1: Two monophyletic clades ---
clade1 <- rtree(4)
clade2 <- rtree(6)

clade1$tip.label <- paste0("Nearctic_", seq_along(clade1$tip.label))
clade2$tip.label <- paste0("Palearctic_", seq_along(clade2$tip.label))

phy_mono <- bind.tree(clade1, clade2, where="root")
cols_mono <- get_colors(phy_mono$tip.label)

# --- Case 2: Random labeling ---
phy_rand <- rtree(10)
regions <- sample(c("Nearctic", "Palearctic"), size=10, replace=TRUE)
phy_rand$tip.label <- paste0(regions, "_", seq_len(10))
cols_rand <- get_colors(phy_rand$tip.label)

# --- Plot side-by-side cladograms ---
par(mfrow=c(1,2), mar=c(2,1,3,1))
plot(phy_mono, tip.color=cols_mono, cex=1, type="cladogram")
title("A")
plot(phy_rand, tip.color=cols_rand, cex=1, type="cladogram")
title("B")
par(mfrow=c(1,1))

```

# Nearctic: 1 endemic Subfamily 

Endemism / endemic: the state of a taxon being found only in a single defined geographic location

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/wrentit.png}
\end{figure}  

Timaliinae (Wrentit)

\hfill \tiny ([*Chamaea fasciata*](https://macaulaylibrary.org/asset/99998731))

# Palearctic: 1 endemic family

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/accentor.png}
\end{figure}  

Prunellidae (accentors)

\hfill \tiny ([*Prunella collaris*](https://macaulaylibrary.org/asset/84441901))

# "Holarctic": 4 endemic families or subfamilies

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/bowa.png}
\end{figure}  

Gaviidae (divers or loons), Tetraoninae (grouse), Alcidae (auks), and Bombycillidae (waxwings).

\hfill \tiny ([*Bombycilla garrulus*](https://macaulaylibrary.org/asset/134340871))

# Neotropical: 31 families

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/rhea.png}
\end{figure}  

Rheidae (Rheas), Ramphastidae (toucans), Jacamars (Galbulidae), many more

\hfill \tiny ([*Rhea americana*](https://macaulaylibrary.org/asset/72353001))

# Afrotropical: 6 families 

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/mousebird.png}
\end{figure}  

Coliidae (mousebirds), Stuthionidae (ostriches), Numididae (guinea fowl)

\hfill \tiny ([*Colius striatus*](https://macaulaylibrary.org/asset/211674951))

# Indomalayan: 1 endemic family 

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/bluebird.png}
\end{figure}  

Irenidae (fairy-bluebirds)

\hfill \tiny ([*Irena tweeddalii*](https://macaulaylibrary.org/asset/574288261))

# Australasian: 13 endemic families 

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/bop.png}
\end{figure}  

Meliphagidae (honeyeaters), Paradisaeidae (birds of paradise), Cacatuidae (cockatoos)

\hfill \tiny ([*Pteridophora alberti*](https://macaulaylibrary.org/asset/106931681))

# Speciation, extinction, and dispersal shape biogeography

Where would you expect to find higher levels of endemism and why?

\begin{figure}
\includegraphics[width = 0.75\linewidth]{images/mtaus.png}
\end{figure}  

a) Montana: It is more centrally located in its continent  
b) Australia: It experiences less extinction  
c) Montana: It experiences more extinction  
d) Australia: It is more isolated from other landmasses  

# References

- Title Slide Species: *Polioptila maior*, *Zentrygon carrikeri*, *Pica pica*, *Spizaetus melanoleucus*  

- Sharpe, R. B. (1909). A hand-list of the genera and species of birds (Nomenclator Avium tum Fossilium tum Viventium) (Vol. V). London: Trustees of the British Museum (Natural History).

- Mayr, E., & Amadon, D. (1951). A classification of recent birds. American Museum Novitates, 1496, 1–42.

- Sibley, C. G., & Monroe, B. L., Jr. (1990). Distribution and taxonomy of birds of the world. New Haven, CT: Yale University Press.

# References

- Schulenberg, T. S., Iliff, M. J., & Clements, J. F. (Eds.). (2014). The eBird/Clements checklist of birds of the world: Version 6.9 [Data set]. Cornell Lab of Ornithology.

- Clements, J. F., Rasmussen, P. C., Schulenberg, T. S., Iliff, M. J., Fredericks, T. A., Gerbracht, J. A., Lepage, D., Spencer, A., Billerman, S. M., Sullivan, B. L., Smith, M., & Wood, C. L. (2024). The eBird/Clements checklist of birds of the world: v2024 [Data set]. Cornell Lab of Ornithology.

- Stiller, J., Feng, S., Chowdhury, A. A., Rivas-González, I., Duchêne, D. A., Fang, Q., ... & Zhang, G. (2024). Complexity of avian evolution revealed by family-level genomes. Nature, 629(8013), 851-860.

- Udvardy, M. D. (1975). A classification of the biogeographical provinces of the world.



